Tensor denoising and completion based on ordinal observations

### 2020-07-30 paper/Camera ready/paper.tex
update

\begin{table*}[ht]
\centerline{\includegraphics[width=1\textwidth]{compare.pdf}}
\caption{Comparison with previous work. For ease of presentation, we summarize the error rate and sample complexity assuming equal tensor dimension in all modes. $K$: tensor order; $L$: number of ordinal levels; $d$: dimension at each mode. }\label{tab:compare}
\end{table*}

@@

We address the challenge by proposing a cumulative link model that enjoys palindromic invariance~\cite{mccullagh1980regression}. A distinctive non-monotonic, phase-transition pattern is demonstrated (Fig~\ref{fig:finite}b). We prove that recovery from quantized tensors achieves equally good information-theoretical convergence as the continuous tensors. These results fills the gap between classical and non-classical (ordinal) observations, thereby greatly enriching the tensor-model literature.

@@

\section{Preliminaries}
Let $\tY\in\mathbb{R}^{d_1\times \cdots \times d_K}$ denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional tensor. We use $y_\omega$ to denote the tensor entry indexed by $\omega$, where $\omega\in[d_1]\times\cdots\times[d_K]$.  The Frobenius norm of $\tY$ is defined as $\FnormSize{}{\tY}=\sum_\omega y^2_\omega$ and the infinity norm is defined as $\mnormSize{}{\tY}=\max_{\omega}|y_\omega|$. We use $\tY_{(k)}$ to denote the unfolded matrix of size $d_k$-by-$\prod_{i\neq k}d_k$, obtained by reshaping the tensor along the mode-$k$. The Tucker rank of $\tY$ is defined as a length-$K$ vector $\mr = (r_1,\ldots,r_K)$, where $r_k$ is the rank of matrix $\tY_{(k)}$ for $k \in[K]$.

@@

\section{Model formulation and motivation}
\subsection{Observation model}
Let $\tY$ denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional data tensor. Suppose the entries of $\tY$ are ordinal-valued, and the observation space consists of $L$ ordered levels, denoted by $[L]:=\{1,\ldots,L\}$. We propose a cumulative link model for the ordinal tensor $\tY=\entry{y_\omega}\in[L]^{d_1\times \cdots\times d_K}$. Specifically, assume the entries $y_\omega$ are (conditionally) independently distributed with cumulative probabilities:
\begin{equation}\label{eq:model}
\mathbb{P}(y_\omega\leq \ell)=f(b_\ell-\theta_\omega),\ \text{for all}\ \ell\in[L-1],
\end{equation}
where $\mb=(b_1,\ldots,b_{L-1})$ is a set of unknown scalars satisfying $b_1<\cdots <b_{L-1}$, $\Theta=\entry{\theta_\omega}\in\mathbb{R}^{d_1\times \cdots \times d_K}$ is a continuous-valued parameter tensor satisfying certain low-dimensional structure (to be specified later), and $f(\cdot)\colon \mathbb{R}\mapsto[0,1]$ is a known, strictly increasing function. We refer to $\mb$ as the cut-off points and $f$ the link function.

The formulation~\eqref{eq:model} imposes an additive model to the transformed probability of cumulative categories. This modeling choice is to respect the ordering structure among the categories. For example, if we choose the inverse link $f^{-1}(x)=\log {x\over 1-x}$ to be the log odds, then the model~\eqref{eq:model} implies linear spacing between the proportional odds:
\begin{equation}\label{eq:logodd}
\log {\mathbb{P}(y_{\omega}\leq \ell) \over \mathbb{P}(y_\omega >  \ell) } - \log {\mathbb{P}(y_{\omega}\leq {\ell-1}) \over \mathbb{P}(y_\omega >  {\ell-1}) } = b_\ell-b_{\ell-1},
\end{equation}
for all tensor entries $y_\omega$. When there are only two categories in the observation space (e.g. binary tensors), the cumulative model~\eqref{eq:model} is equivalent to the usual binomial link model. In general, however, when the number of categories $L\geq 3$, the proportional odds assumption~\eqref{eq:logodd} is more parsimonious, in that, the ordered categories can be envisaged as contiguous intervals on the continuous scale, where the points of division are exactly $b_1<\cdots <b_{L-1}$. This interpretation will be made more explicit in the next section.

@@

The ordinal tensor model~\eqref{eq:model} with certain types of link $f$ has the equivalent representation as an $L$-level quantization model on $\tY=\entry{y_\omega}$:
\begin{align}\label{eq:quantization}
%y_\omega=\sum_{\ell\in[L]} \ell \mathds{1}\{y^*_\omega \in(b_{\ell-1},b_\ell]\},\ \forall \ \omega\in[d_1]\times \cdots \times [d_k],
%\end{equation}
y_\omega&=
\begin{cases}
1,& \text{if $y^*_\omega\in(-\infty, b_1]$},\\
2,& \text{if $y^*_\omega\in(b_1, b_2]$},\\
\vdots  &\vdots\\
L,& \text{if $y^*_\omega\in(b_{L-1}, \infty)$},\\
\end{cases}
\end{align}
for all $\omega\in[d_1]\times \cdots \times [d_k]$. Here, $\tY^*=\entry{y^*_\omega}$ is a latent continuous-valued tensor following an additive noise model:
\begin{align}\label{eq:latent}
\KeepStyleUnderBrace{\tY^*}_{\text{latent continuous-valued tensor}}=\KeepStyleUnderBrace{\Theta}_{\text{signal tensor}}+\KeepStyleUnderBrace{\tE}_{\text{i.i.d.\ noise}},
\end{align}
where $\tE=\entry{\varepsilon_\omega}\in\mathbb{R}^{d_1\times \cdots \times d_K}$ is a noise tensor with i.i.d.\ entries according to distribution $\mathbb{P}(\varepsilon)$. From the viewpoint of~\eqref{eq:latent}, the parameter tensor $\Theta$ can be interpreted as the latent signal tensor prior to contamination and quantization.

@@

\begin{example}[Logistic model] The logistic model is characterized by~\eqref{eq:model} with $f(\theta)=(1+e^{-\theta/\sigma})^{-1}$, where $\sigma>0$ is the scale parameter. Equivalently, the noise $\varepsilon_\omega$ in~\eqref{eq:quantization} follows i.i.d.\ logistic distribution with scale parameter $\sigma$.
\end{example}
\begin{example}[Probit model] The probit model is characterized by~\eqref{eq:model} with
$f(\theta)=\mathbb{P}(z\leq \theta/\sigma)$, where $z\sim N(0,1)$. Equivalently, the noise $\varepsilon_\omega$ in~\eqref{eq:quantization} follows i.i.d.\ $N(0,\sigma^2)$.
\end{example}
Other link functions are also possible, such as Laplace, Cauchy, etc~\cite{mccullagh1980regression}. All the models share the property that the ordered categories can be thought of as contiguous interval on some continuous scale. We should point out that, although the latent-variable interpretation is incisive, our estimation procedure does not refer to the existence of $\tY^*$. Therefore, our model~\eqref{eq:model} is general and still valid in the absence of quantization process. More generally, we make the following assumptions about the link $f$.

@@

Clearly, the problem (P1) cannot be solved uniformly for all possible $\Theta$. We focus on a class of ``low-rank'' and ``flat'' signal tensors, which is a plausible assumption in practical applications~\cite{zhou2013tensor,bhaskar20151}. Specifically, we consider the parameter space:
\begin{equation}\label{eq:space}
\tP=\left\{\Theta\in\mathbb{R}^{d_1\times \cdots \times d_K} \colon \text{rank}(\Theta)\leq \mr,\ \mnormSize{}{\Theta}\leq \alpha\right\}.
\end{equation}
where $\mr=(r_1,\ldots,r_K)$ denotes the Tucker rank of $\Theta$.

The parameter tensor of our interest satisfies two constraints. The first is that $\Theta$ is a low-rank tensor, with $r_k=\tO(1)$ for all $k\in[K]$. Equivalently, $\Theta$ admits the Tucker decomposition:
\begin{equation}\label{eq:Tucker}
\Theta=\tC\times_1\mM_1\times_1 \cdots \times_K \mM_K,
\end{equation}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a core tensor, $\mM_k\in\mathbb{R}^{d_k\times r_k}$ are factor matrices with orthogonal columns, and $\times_k$ denotes the tensor-by-matrix multiplication~\cite{kolda2009tensor}. The Tucker low-rankness is popularly imposed in tensor data analysis, and is shown to provide a reasonable tradeoff between model complexity and model flexibility. Note that, unlike matrices, there are various notations of tensor low-rankness, such as CP rank~\cite{hitchcock1927expression} and train rank~\cite{oseledets2011tensor}. Some notation of low-rankness may lead to mathematically ill-posed optimization; for example, the best low CP-rank tensor approximation may not exist~\cite{de2008tensor}. We choose Tucker representation for well-posedness of optimization and easy interpretation.

The second constraint is that the entries of $\Theta$ are uniformly bounded in magnitude by a constant $\alpha \in \mathbb{R}_{+}$. In view of~\eqref{eq:latent}, we refer to $\alpha$ as the signal level. The entry-wise bound assumption is a technical condition that avoids the degeneracy in probability estimation with ordinal observations.

\subsection{Problem 2: Tensor completion}
Motivated by applications in collaborative filtering, we also consider a more general setup when only a subset of tensor entries $y_\omega$ are observed. Let $\Omega\subset[d_1]\times \cdots\times[d_K]$ denote the set of observed indices. The second question we aim to address is stated as follows:

@@

\section{Rank-constrained M-estimator}\label{sec:theory}
We present a general treatment to both problems mentioned above. With a little abuse of notation, we use $\Omega$ to denote either the full index set $\Omega=[d_1]\times \cdots \times [d_K]$ (for the tensor denoising) or a random subset induced from the sampling distribution $\Pi$ (for the tensor completion). Define $b_0=-\infty$, $b_L=\infty$, $f(-\infty)=0$ and $f(\infty)=1$. The log-likelihood associated with the observed entries is
\begin{align}\label{eq:objective}
\logl(\Theta, \mb)&=\sum_{\omega\in\Omega}\sum_{\ell\in[L]} \Big\{\mathds{1}\{y_\omega=\ell\} \log \big[f(b_\ell-\theta_\omega)- \notag \\
&\quad \quad \quad \quad \quad \quad \quad  f(b_{\ell-1}-\theta_\omega)\big]\Big\}.
\end{align}
We propose a rank-constrained maximum likelihood estimator (a.k.a.\ M-estimator) for $\Theta$:
\begin{align}\label{eq:estimator}
\hat \Theta&=\argmax_{\Theta\in \tP}\logl(\Theta, \mb),\ \text{where}\notag \\
\tP&=\left\{\Theta\in\mathbb{R}^{d_1\times \cdots \times d_K} \colon \text{rank}(\Theta)\leq \mr,\ \mnormSize{}{\Theta}\leq \alpha\right\}.
 \end{align}
In practice, the cut-off points $\mb$ are unknown and should be jointly estimated with $\Theta$. We will focus on the theory and algorithm with known $\mb$ in the main paper. The adaptation for unknown $\mb$ is addressed in Section~\ref{sec:algorithm} and the Supplement.

We define a few key quantities that will be used in our theory. Let $g_\ell=f(\theta+b_\ell)-f(\theta+b_{\ell-1})$ for all $\ell\in[L]$, and
\begin{align}\label{eq:regular}
A_\alpha &= \min_{\ell\in[L], |\theta|\leq \alpha} g_\ell(\theta),\quad  U_\alpha=\max_{\ell\in[L], |\theta|\leq \alpha } {|\dot{g}_\ell(\theta)|\over g_\ell(\theta)},\\
 L_\alpha &= \min_{\ell\in[L], |\theta|\leq \alpha }\left[{\dot{g}_\ell^2 (\theta)\over g_\ell^2(\theta)} -{\ddot{g}_\ell(\theta)\over g_\ell(\theta)}\right],
\end{align}
where $\dot{g}(\theta)=dg(\theta)/d\theta$, and $\alpha$ is the entrywise bound of $\Theta$. In view of equation~\eqref{eq:latent}, these quantities characterize the geometry including flatness and convexity of the latent noise distribution. Under the Assumption~\ref{ass:link}, all these quantities are strictly positive and independent of tensor dimension.

@@

\begin{thm}[Statistical convergence] \label{thm:rate}
Consider an ordinal tensor $\tY\in[L]^{d_1\times\dots\times d_K}$ generated from model~\eqref{eq:model}, with the link function $f$ and the true coefficient tensor $\trueT\in\tP$. Define $r_{\max}=\max_k r_k$. Then, with very high probability, the estimator in~\eqref{eq:estimator} satisfies
\begin{equation}\label{eq:rate}
\mathrm{MSE}(\hat \Theta, \trueT) \leq \min\left( 4\alpha^2,\ {c_1  U^2_\alpha r_{\max}^{K-1}  \over  L^2_\alpha } {\sum_kd_k\over  \prod_k d_k} \right),
\end{equation}
where $c_1 >0$ is a constant that depends only on $K$.
\end{thm}
Theorem~\ref{thm:rate} establishes the statistical convergence for the estimator~\eqref{eq:estimator}. In fact, the proof of this theorem (see the Supplement) shows that the same statistical rate holds, not only for the global optimizer~\eqref{eq:estimator}, but also for any local optimizer $\check \Theta$ in the level set $\{\check\Theta\in\tP: \logl(\check\Theta)\geq \logl(\trueT)\}$. This suggests that the local optimality itself is not necessarily a severe concern in our context, as long as the convergent objective is large enough. In Section ~\ref{sec:algorithm}, we perform empirical studies to assess the algorithmic stability.

A similar conclusion is obtained for the prediction error, measured in Kullback-Leibler (KL) divergence, between the categorical distributions in the observation space.
\begin{cor}[Prediction error]~\label{cor:prediction}
Assume the same set-up as in Theorem~\ref{thm:rate}. Let $\mathbb{P}_{\tY}$ and $\hat{\mathbb{P}}_{\tY}$ denote the distributions generating the $L$-level ordinal tensor $\tY$, given the true parameter $\Theta$ and its estimator $\hat \Theta$, respectively. Assume $L\geq 2$. Then, with very high probability,
\begin{equation}\label{eq:KLrate}
\text{KL}(\mathbb{P}_{\tY} || \hat{\mathbb{P}}_{\tY}) \leq  {c_1 U^2_\alpha r_{\max}^{K-1}  \over L^2_\alpha } {(4L-6)\dot{f}^2(0)  \over A_\alpha} {\sum_kd_k\over  \prod_k d_k},
\end{equation}
where $c_1 >0$ is the same constant as in Theorem~\ref{thm:rate}.
\end{cor}
To gain insight into these bounds, we consider a special setting with equal dimension in all modes, i.e., $d_1=\cdots=d_K=d$. In such a case, our bound \eqref{eq:rate} reduces to
\begin{equation}\label{eq:ours}
\text{MSE}(\hat \Theta, \trueT) \asymp d^{-(K-1)}, \quad \text{as}\ d\to \infty.
\end{equation}
Hence, our estimator achieves consistency with polynomial convergence rate. We compare the bound with existing literature. In the special case $L=2$, \citet{ghadermarzy2018learning} proposed a max-norm constrained estimator $\tilde{\Theta}$ with $\text{MSE}(\tilde{\Theta}, \trueT) \asymp  d^{-(K-1)/2}$. In contrast, our estimator converges at a rate of $d^{-(K-1)}$, which is substantially faster than theirs. This provides a positive answer to the open question posed in~\citet{ghadermarzy2018learning} whether the square root in the bound is removable. The improvement stems from utilizing the exact low-rankness of $\Theta$, whereas the surrogate rank measure employed in~\citet{ghadermarzy2018learning} is scale-sensitive.

Our bound also generalizes the previous results on ordinal matrices. The convergence rate for rank-constrained matrix estimation was $\tO(1/\sqrt{d})$~\citep{bhaskar2016probabilistic}, which fits into our special case when $K=2$. Furthermore, our results~\eqref{eq:rate} and~\eqref{eq:KLrate} reveal that the convergence becomes favorable as the order of data tensor increases. Intuitively, the sample size for tensor data analysis is the number of entries, $\prod_k d_k$, and the number of free parameters is roughly on the order of $\sum_{k}d_k$, assuming $r_{\max}=\tO(1)$. A higher tensor order implies higher effective sample size per parameter, and thus exhibits a faster convergence rate in high dimensions.

We next show the statistical optimality of our estimator $\hat \Theta$. The result is based on the information theory, and applies to all estimators in $\tP$, including but not limited to $\hat \Theta$ in~\eqref{eq:estimator}.

@@

\subsection{Sample complexity for tensor completion}
We now consider the tensor completion problem, when only a subset of entries $\Omega$ are observed. We consider a general sampling procedure induced by $\Pi$. The recovery accuracy is assessed by the weighted squared error:
\begin{align}\label{eq:weighted}
\PiFnormSize{}{\Theta-\hat \Theta}^2&\stackrel{\text{def}}{=}
{1\over |\Omega|}\mathbb{E}_{\Omega\sim \Pi}\FnormSize{}{\Theta-\hat \Theta}^2\notag \\
&=\sum_{\omega\in[d_1]\times \cdots \times [d_K]} \pi_{\omega}(\Theta_{\omega}-\hat \Theta_{\omega})^2.
\end{align}

Note that the recovery error depends on the distribution $\Pi$. In particular, tensor entries with higher sampling probabilities have more influence on the recovery accuracy, compared to the ones with lower sampling probabilities.
\begin{rmk} If we assume each entry is sampled with strictly positive probability; i.e.\ there exits a constant $\mu> 0$ s.t.
\[
\ \pi_\omega\geq {1\over \mu \prod_k d_k},\quad \text{for all}\ \omega\in[d_1]\times\cdots \times [d_K],
\]
then the error in~\eqref{eq:weighted} provides an upper bound for MSE:
\[
\PiFnormSize{}{\Theta-\hat \Theta}^2 \geq {\FnormSize{}{\Theta-\hat \Theta}^2\over \mu \prod_kd_k}={1\over \mu}\text{MSE}(\hat \Theta, \trueT).
\]
The equality is attained under uniform sampling with $\mu=1$.
\end{rmk}

>>>>

\begin{table*}[ht]
\centerline{\includegraphics[width=1\textwidth]{compare.pdf}}
\caption{Comparison with previous work. For ease of presentation, we summarize the error rate and sample complexity assuming equal tensor dimension in all modes. $K$: tensor order; $L$: number of ordinal levels; $d$: dimension at each mode. }\label{tab:compare}
\vspace{-.2cm}
\end{table*}

@@

We address the challenge by proposing a cumulative link model that enjoys palindromic invariance~\cite{mccullagh1980regression}. A distinctive non-monotonic, phase-transition pattern is demonstrated, as we show in Section~\ref{sec:experiment}. We prove that the recovery from quantized tensors achieves equally good information-theoretical convergence as the continuous tensors. These results fills the gap between classical and non-classical (ordinal) observations, thereby greatly enriching the tensor model literature.

@@

\section{Preliminaries}
Let $\tY\in\mathbb{R}^{d_1\times \cdots \times d_K}$ denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional tensor. We use $y_\omega$ to denote the tensor entry indexed by $\omega$, where $\omega\in[d_1]\times\cdots\times[d_K]$.  The Frobenius norm of $\tY$ is defined as $\FnormSize{}{\tY}=\sum_\omega y^2_\omega$ and the infinity norm is defined as $\mnormSize{}{\tY}=\max_{\omega}|y_\omega|$. We use $\tY_{(k)}$ to denote the unfolded matrix of size $d_k$-by-$\prod_{i\neq k}d_i$, obtained by reshaping the tensor along the mode $k\in[K]$. The Tucker rank of $\tY$ is defined as a length-$K$ vector $\mr = (r_1,\ldots,r_K)$, where $r_k$ is the rank of matrix $\tY_{(k)}$ for $k \in[K]$.

@@

\section{Model formulation and motivation}
\subsection{Observation model}
Let $\tY$ denote an order-$K$ $(d_1,\ldots,d_K)$-dimensional data tensor. Suppose the entries of $\tY$ are ordinal-valued, and the observation space consists of $L$ ordered levels, denoted by $[L]=\{1,\ldots,L\}$. We propose a cumulative link model for the ordinal tensor $\tY=\entry{y_\omega}\in[L]^{d_1\times \cdots\times d_K}$. Specifically, assume the entries $y_\omega$ are (conditionally) independently distributed with cumulative probabilities,
\begin{equation}\label{eq:model}
\mathbb{P}(y_\omega\leq \ell)=f(b_\ell-\theta_\omega),\ \text{for all}\ \ell\in[L-1],
\end{equation}
where $\mb=(b_1,\ldots,b_{L-1})$ is a set of unknown scalars satisfying $b_1<\cdots <b_{L-1}$, $\Theta=\entry{\theta_\omega}\in\mathbb{R}^{d_1\times \cdots \times d_K}$ is a continuous-valued parameter tensor satisfying certain low-dimensional structure (to be specified later), and $f(\cdot)\colon \mathbb{R}\mapsto[0,1]$ is a known, strictly increasing function. We refer to $\mb$ as the cut-off points and $f$ the link function.

The formulation~\eqref{eq:model} imposes an additive model to the transformed probability of cumulative categories. This modeling choice is to respect the ordering structure among the categories. For example, if we choose the inverse link $f^{-1}(x)=\log {x\over 1-x}$ to be the log odds, then the model~\eqref{eq:model} implies a linear spacing between the proportional odds,
\begin{equation}\label{eq:logodd}
\log {\mathbb{P}(y_{\omega}\leq \ell) \over \mathbb{P}(y_\omega >  \ell) } - \log {\mathbb{P}(y_{\omega}\leq {\ell-1}) \over \mathbb{P}(y_\omega >  {\ell-1}) } = b_\ell-b_{\ell-1},
\end{equation}
for all tensor entries $y_\omega$. When there are only two categories in the observation space (e.g., for binary tensors), the cumulative model~\eqref{eq:model} is equivalent to the usual binomial link model. In general, however, when the number of categories $L\geq 3$, the proportional odds assumption~\eqref{eq:logodd} is more parsimonious, in that, the ordered categories can be envisaged as contiguous intervals on the continuous scale, where the points of division are exactly $b_1<\cdots <b_{L-1}$. This interpretation will be made explicit in the next section.

@@

The ordinal tensor model~\eqref{eq:model} with certain types of link $f$ has the equivalent representation as an $L$-level quantization model on $\tY=\entry{y_\omega}$:
\begin{align}\label{eq:quantization}
%y_\omega=\sum_{\ell\in[L]} \ell \mathds{1}\{y^*_\omega \in(b_{\ell-1},b_\ell]\},\ \forall \ \omega\in[d_1]\times \cdots \times [d_k],
%\end{equation}
y_\omega&=
\begin{cases}
1,& \text{if $y^*_\omega\in(-\infty, b_1]$},\\
2,& \text{if $y^*_\omega\in(b_1, b_2]$},\\
\vdots  &\vdots\\
L,& \text{if $y^*_\omega\in(b_{L-1}, \infty)$},\\
\end{cases}
\end{align}
for all $\omega\in[d_1]\times \cdots \times [d_k]$. Here, $\tY^*=\entry{y^*_\omega}$ is a latent continuous-valued tensor following an additive noise model,
\begin{align}\label{eq:latent}
\KeepStyleUnderBrace{\tY^*}_{\text{latent continuous-valued tensor}}=\KeepStyleUnderBrace{\Theta}_{\text{signal tensor}}+\KeepStyleUnderBrace{\tE}_{\text{i.i.d.\ noise}},
\end{align}
where $\tE=\entry{\varepsilon_\omega}\in\mathbb{R}^{d_1\times \cdots \times d_K}$ is a noise tensor with independent and identically distributed (i.i.d.) entries according to distribution $\mathbb{P}(\varepsilon)$. From the viewpoint of~\eqref{eq:latent}, the parameter tensor $\Theta$ can be interpreted as the latent signal tensor prior to contamination and quantization.

@@

\begin{example}[Logistic model] The logistic model is characterized by~\eqref{eq:model} with $f(\theta)=(1+e^{-\theta/\sigma})^{-1}$, where $\sigma>0$ is the scale parameter. Equivalently, the noise $\varepsilon_\omega$ in~\eqref{eq:quantization} follows i.i.d.\ logistic distribution with scale parameter $\sigma$.
\end{example}
\begin{example}[Probit model] The probit model is characterized by~\eqref{eq:model} with
$f(\theta)=\mathbb{P}(z\leq \theta/\sigma)$, where $z\sim N(0,1)$. Equivalently, the noise $\varepsilon_\omega$ in~\eqref{eq:quantization} follows i.i.d.\ $N(0,\sigma^2)$.
\end{example}
Other link functions are also possible, such as Laplace, Cauchy, etc~\cite{mccullagh1980regression}. These latent variable models share the property that the ordered categories can be thought of as contiguous intervals on some continuous scale. We should point out that, although the latent-variable interpretation is incisive, our estimation procedure does not refer to the existence of $\tY^*$. Therefore, our model~\eqref{eq:model} is general and still valid in the absence of quantization process. More generally, we make the following assumptions about the link $f$.

@@

Clearly, the problem (P1) cannot be solved uniformly for all possible $\Theta$ with no assumptions. We focus on a class of ``low-rank'' and ``flat'' signal tensors, which is a plausible assumption in practical applications~\cite{zhou2013tensor,bhaskar20151}. Specifically, we consider the parameter space,
\begin{equation}\label{eq:space}
\tP=\left\{\Theta\in\mathbb{R}^{d_1\times \cdots \times d_K} \colon \text{rank}(\Theta)\leq \mr,\ \mnormSize{}{\Theta}\leq \alpha\right\},
\end{equation}
where $\mr=(r_1,\ldots,r_K)$ denotes the Tucker rank of $\Theta$.

The parameter tensor of our interest satisfies two constraints. The first is that $\Theta$ is a low-rank tensor, with $r_k=\tO(1)$ as $d_{\min}\to \infty$ for all $k\in[K]$. Equivalently, $\Theta$ admits the Tucker decomposition:
\begin{equation}\label{eq:Tucker}
\Theta=\tC\times_1\mM_1\times_1 \cdots \times_K \mM_K,
\end{equation}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a core tensor, $\mM_k\in\mathbb{R}^{d_k\times r_k}$ are factor matrices with orthogonal columns, and $\times_k$ denotes the tensor-by-matrix multiplication~\cite{kolda2009tensor}. The Tucker low-rankness is popularly imposed in tensor analysis, and the rank determines the tradeoff between model complexity and model flexibility. Note that, unlike matrices, there are various notions of tensor low-rankness, such as CP rank~\cite{hitchcock1927expression} and train rank~\cite{oseledets2011tensor}. Some notions of low-rankness may lead to mathematically ill-posed optimization; for example, the best low CP-rank tensor approximation may not exist~\cite{de2008tensor}. We choose Tucker representation for well-posedness of optimization and easy interpretation.

The second constraint is that the entries of $\Theta$ are uniformly bounded in magnitude by a constant $\alpha \in \mathbb{R}_{+}$. In view of~\eqref{eq:latent}, we refer to $\alpha$ as the signal level. The boundedness assumption is a technical condition that avoids the degeneracy in probability estimation with ordinal observations.
\vspace{-.2cm}
\subsection{Problem 2: Tensor completion}
Motivated by applications in collaborative filtering, we also consider a more general setup when only a subset of tensor entries $y_\omega$ are observed. Let $\Omega\subset[d_1]\times \cdots\times[d_K]$ denote the set of observed indices. The second question we aim to address is stated as follows:

@@

\section{Rank-constrained M-estimator}\label{sec:theory}
We present a general treatment to both problems mentioned above. With a little abuse of notation, we use $\Omega$ to denote either the full index set $\Omega=[d_1]\times \cdots \times [d_K]$ (for the tensor denoising) or a random subset induced from the sampling distribution $\Pi$ (for the tensor completion). Define $b_0=-\infty$, $b_L=\infty$, $f(-\infty)=0$ and $f(\infty)=1$. The log-likelihood associated with the observed entries is
\begin{align}\label{eq:objective}
\logl(\Theta, \mb)&=\sum_{\omega\in\Omega}\sum_{\ell\in[L]} \Big\{\mathds{1}\{y_\omega=\ell\} \log \big[f(b_\ell-\theta_\omega)- \notag \\
&\quad \quad \quad \quad \quad \quad \quad  f(b_{\ell-1}-\theta_\omega)\big]\Big\}.
\end{align}
We propose a rank-constrained maximum likelihood estimator (a.k.a.\ M-estimator) for $\Theta$,
\begin{align}\label{eq:estimator}
\hat \Theta&=\argmax_{\Theta\in \tP}\logl(\Theta, \mb),\ \text{where}\notag \\
\tP&=\left\{\Theta\in\mathbb{R}^{d_1\times \cdots \times d_K} \colon \text{rank}(\Theta)\leq \mr,\ \mnormSize{}{\Theta}\leq \alpha\right\}.
 \end{align}
In practice, the cut-off points $\mb$ are unknown and should be jointly estimated with $\Theta$. We will present the theory and algorithm with known $\mb$ in the main paper. The adaptation for unknown $\mb$ is addressed in Section~\ref{sec:algorithm} and the Supplement.

We define a few key quantities that will be used in our theory. Let $g_\ell=f(\theta+b_\ell)-f(\theta+b_{\ell-1})$ for all $\ell\in[L]$, and
\begin{align}\label{eq:regular}
A_\alpha &= \min_{\ell\in[L], |\theta|\leq \alpha} g_\ell(\theta),\quad  U_\alpha=\max_{\ell\in[L], |\theta|\leq \alpha } {|\dot{g}_\ell(\theta)|\over g_\ell(\theta)},\\
 L_\alpha &= \min_{\ell\in[L], |\theta|\leq \alpha }\left[{\dot{g}_\ell^2 (\theta)\over g_\ell^2(\theta)} -{\ddot{g}_\ell(\theta)\over g_\ell(\theta)}\right],
\end{align}
where $\dot{g}(\theta)=dg(\theta)/d\theta$, and $\alpha$ is the entrywise bound of $\Theta$. In view of equation~\eqref{eq:latent}, these quantities characterize the geometry including flatness and convexity of the latent noise distribution. Under Assumption~\ref{ass:link}, all these quantities are strictly positive and independent of tensor dimension.

@@

\begin{thm}[Statistical convergence] \label{thm:rate}
Consider an ordinal tensor $\tY\in[L]^{d_1\times\dots\times d_K}$ generated from model~\eqref{eq:model}, with the link function $f$ and the true coefficient tensor $\trueT\in\tP$. Define $r_{\max}=\max_k r_k$. Then, with very high probability, the estimator in~\eqref{eq:estimator} satisfies
\begin{equation}\label{eq:rate}
\mathrm{MSE}(\hat \Theta, \trueT) \leq \min\left( 4\alpha^2,\ {c_1  U^2_\alpha r_{\max}^{K-1}  \over  L^2_\alpha } {\sum_kd_k\over  \prod_k d_k} \right),
\end{equation}
where $c_1 >0$ is a constant that depends only on $K$.
\end{thm}
Theorem~\ref{thm:rate} establishes the statistical convergence for the estimator~\eqref{eq:estimator}. In fact, the proof of this theorem (see the Supplement) shows that the same statistical rate holds, not only for the global optimizer~\eqref{eq:estimator}, but also for any local optimizer $\check \Theta$ in the level set $\{\check\Theta\in\tP\colon \logl(\check\Theta)\geq \logl(\trueT)\}$. This suggests that the local optimality itself is not necessarily a severe concern in our context, as long as the convergent objective is large enough. In Section ~\ref{sec:algorithm}, we perform empirical studies to assess the algorithmic stability.

To gain insight into the bound~\eqref{eq:rate}, we consider a special setting with equal dimension in all modes, i.e., $d_1=\cdots=d_K=d$. In such a case, our bound \eqref{eq:rate} reduces to
\begin{equation}\label{eq:ours}
\text{MSE}(\hat \Theta, \trueT) \asymp d^{-(K-1)}, \quad \text{as}\ d\to \infty.
\end{equation}
Hence, our estimator achieves consistency with polynomial convergence rate. We compare the bound with existing literature. In the special case $L=2$, \citet{ghadermarzy2018learning} proposed a max-norm constrained estimator $\tilde{\Theta}$ with $\text{MSE}(\tilde{\Theta}, \trueT) \asymp  d^{-(K-1)/2}$. In contrast, our estimator converges at a rate of $d^{-(K-1)}$, which is substantially faster than theirs. This provides a positive answer to the open question posed in~\citet{ghadermarzy2018learning} whether the square root in the bound is removable. The improvement stems from the fact that we have used the exact low-rankness of $\Theta$, whereas the surrogate rank measure employed in~\citet{ghadermarzy2018learning} is scale-sensitive.

Our bound also generalizes the previous results on ordinal matrices. The convergence rate for rank-constrained matrix estimation is $\tO(1/\sqrt{d})$~\citep{bhaskar2016probabilistic}, which fits into our special case when $K=2$. Furthermore, our result~\eqref{eq:rate} reveals that the convergence becomes favorable as the order of data tensor increases. Intuitively, the sample size for analyzing a data tensor is the number of entries, $\prod_k d_k$, and the number of free parameters is roughly on the order of $\sum_{k}d_k$, assuming $r_{\max}=\tO(1)$. A higher tensor order implies higher effective sample size per parameter, thus achieving a faster convergence rate in high dimensions.

A similar conclusion is obtained for the prediction error, measured in Kullback-Leibler (KL) divergence, between the categorical distributions in the observation space.
\begin{cor}[Prediction error]~\label{cor:prediction}
Assume the same set-up as in Theorem~\ref{thm:rate}. Let $\mathbb{P}_{\tY}$ and $\hat{\mathbb{P}}_{\tY}$ denote the distributions generating the $L$-level ordinal tensor $\tY$, given the true parameter $\Theta$ and its estimator $\hat \Theta$, respectively. Assume $L\geq 2$. Then, with very high probability,
\begin{equation}\label{eq:KLrate}
\text{KL}(\mathbb{P}_{\tY} || \hat{\mathbb{P}}_{\tY}) \leq  {c_1 U^2_\alpha r_{\max}^{K-1}  \over L^2_\alpha } {(4L-6)\dot{f}^2(0)  \over A_\alpha} {\sum_kd_k\over  \prod_k d_k},
\end{equation}
where $c_1 >0$ is the same constant as in Theorem~\ref{thm:rate}.
\end{cor}

We next show the statistical optimality of our estimator $\hat \Theta$. The result is based on the information theory and applies to all estimators in $\tP$, including but not limited to $\hat \Theta$ in~\eqref{eq:estimator}.

@@

\subsection{Sample complexity for tensor completion}
We now consider the tensor completion problem, when only a subset of entries $\Omega$ are observed. We consider a general sampling procedure induced by $\Pi$. The recovery accuracy is assessed by the weighted squared error,
\begin{align}\label{eq:weighted}
\PiFnormSize{}{\Theta-\hat \Theta}^2&\stackrel{\text{def}}{=}
{1\over |\Omega|}\mathbb{E}_{\Omega\sim \Pi}\FnormSize{}{\Theta-\hat \Theta}^2\notag \\
&=\sum_{\omega\in[d_1]\times \cdots \times [d_K]} \pi_{\omega}(\Theta_{\omega}-\hat \Theta_{\omega})^2.
\end{align}

Note that the recovery error depends on the distribution $\Pi$. In particular, tensor entries with higher sampling probabilities have more influence on the recovery accuracy, compared to the ones with lower sampling probabilities.
\begin{rmk} If we assume each entry is sampled with some strictly positive probability; i.e.,\ there exits a constant $\mu> 0$ such that
\[
\ \pi_\omega\geq {1\over \mu \prod_k d_k},\quad \text{for all}\ \omega\in[d_1]\times\cdots \times [d_K],
\]
then the error in~\eqref{eq:weighted} provides an upper bound for MSE:
\[
\PiFnormSize{}{\Theta-\hat \Theta}^2 \geq {\FnormSize{}{\Theta-\hat \Theta}^2\over \mu \prod_kd_k}={1\over \mu}\text{MSE}(\hat \Theta, \trueT).
\]
The equality is attained under uniform sampling with $\mu=1$.
\end{rmk}
